\documentclass[12pt,a4paper,openright,twoside]{book}
\usepackage[utf8]{inputenc}
\usepackage{disi-thesis}
\usepackage{code-lstlistings}
\usepackage{notes}
\usepackage{shortcuts}
\usepackage{acronym}
\usepackage{svg}
\usepackage{multirow}
\usepackage{subcaption}

\school{\unibo}
\programme{Second Cycle Degree in Computer Science and Engineering}
\title{Type-Safe Complex Specifications in Kotlin: a DSL to Configure the Alchemist Simulator}
\author{Marco Frattarola}
\date{\today}
\subject{Software Process Engineering}
\supervisor{Prof. Gianluca Aguzzi}
\cosupervisor{Prof. Danilo Pianini}
\morecosupervisor{}
\session{I}
\academicyear{2024-2025}



\mainlinespacing{1.241} % line spacing in mainmatter, comment to default (1)

\begin{document}

\frontmatter\frontispiece

\begin{abstract}	
Modern software systems rely on declarative configuration to specify complex behaviors, yet the industry has
 converged on untyped formats (XML, JSON, and YAML) that introduce severe limitations: 
 runtime errors, implicit type validation, and limited code reuse mechanisms.

This thesis addresses the challenge of replacing untyped declarative configuration formats with 
type-safe internal Domain-Specific Languages (DSLs) in Kotlin 
for configuring complex systems and behaviors. 
The central challenge is reconciling static type verification with dynamic extensibility requirements 
of frameworks supporting arbitrary class loading and runtime component discovery.

The solution employs generative techniques, leveraging compiler plugin APIs like Kotlin Symbol Processing (KSP), 
to bridge simplified DSL APIs with legacy component interfaces while maintaining type safety and enabling compile-time validation.
The approach demonstrates that transitioning to type-safe configuration languages yields substantial benefits: 
compile-time error detection, comprehensive IDE support, and evolution from monolithic files to composable, reusable modules.

The work is validated through the development of a DSL for Alchemist, a general-purpose simulation framework, 
which maintains semantic equivalence with legacy YAML configurations while providing type safety, tooling support,
 and modularity benefits.
This case study demonstrates how type-safe DSLs can successfully 
replace untyped configuration formats in complex, extensible systems.
\end{abstract}

\begin{dedication} % this is optional
To my family.
\end{dedication}

%----------------------------------------------------------------------------------------
\tableofcontents   
\listoffigures     % (optional) comment if empty
\lstlistoflistings % (optional) comment if empty
%----------------------------------------------------------------------------------------

\mainmatter

\input{chapters/introduction}
\input{chapters/background}
\input{chapters/analysis}
\input{chapters/design}
\input{chapters/implementation}
\input{chapters/evaluation}
\input{chapters/conclusion}

%----------------------------------------------------------------------------------------
% BIBLIOGRAPHY
%----------------------------------------------------------------------------------------

\backmatter

\nocite{*} % Remove this as soon as you have the first citation

\bibliographystyle{unsrt}
\bibliography{bibliography}

\begin{acknowledgements}
I would like to thank my supervisors, Prof. Gianluca Aguzzi and Prof. Danilo Pianini, 
for their guidance and support throughout the development of this thesis. 

My heartfelt thanks go to my family for their encouragement and patience. 
I am particularly grateful to Carlotta for making this journey easier and for standing by me 
over the years, even in the most difficult times.

Last but not least, I would like to acknowledge my cat, Chicco, 
who kept me company during the long working sessions.

\end{acknowledgements}

\end{document}
